Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-10 paper/manuscript.tex
Updates from Overleaf

\begin{figure}
\includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
\caption{\label{fig:overlap}Overlap matrices showing (A) good, (B) mediocre, and (C) poor overlap.}
\end{figure}

Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} A, B, and C respectively. For (A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good. In the case of (B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained. For (C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states. While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\lambda$ windows.

Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of lambda windows. However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time. While variance can be decreased both by increasing simulation length and by increasing the number of lambda values, if overlap is poor, the latter approach often has a larger immediate impact.  Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows. Different approaches are described in section \ref{part_of_section_six}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.

On a final note, it has been argued that overlap matrices are only a qualitative estimator for the spacing of windows. Another estimator named Time Derivative of overall Variance (TDV)~\cite{sun2017barbased} has been suggested for this task, which is an explicit function of the overall variance. However, it could not be proven that better suggestions for the $\lambda$ spacing can be obtained by this approach compared to overlap matrices. Hence, it should rather be seen as a complementary approach.

In a nutshell, it must be ensured that the values of an overlap matrix in the first diagonals above and below the main diagonal are at least 0.03 in order to yield reliable results.

There are several other parameters which should be investigated. Basic analysis of the trajectories should be performed to ensure that the ligand does not tumble out of the binding site by measuring its RMSD. Furthermore, distributions of the ligand's dihedral angles can be obtained and compared to experimental values. Thus, poorly sampled conformations can be identified and the issue can be adequately addressed, e.g. by prolonging the simulation time or choosing a different starting conformation. In general, running the same perturbation using different ligand poses may offer insights into whether the calculated free energies are dependent on the initial conformation.
Another easy check can be performed by comparing the results obtained by different free energy estimators (e.g. TI and MBAR). In principal, the outcomes of both estimators should be the same (within a certain range). Hence, if the outcomes deviate by more than a predefined threshold (e.g. 1.0 kcal/mol), the free energies may be unreliable and further calculations are necessary.
\todo[inline]{MK: @ASJSM: Please review and add a few sentences on replica mixing and correlation time}
\begin{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\item Other considerations for many transformations
\end{itemize}

@@

>>>>

\begin{figure}
\includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
\caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data. (A) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (B) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values. (C)shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
\end{figure}

Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} (A), (B), and (C) respectively. For (A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good. In the case of (B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained. For (C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states. While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\lambda$ windows.
%
Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of lambda windows. However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time. While variance can be decreased both by increasing simulation length and by increasing the number of lambda values, if overlap is poor, the latter approach often has a larger immediate impact.  Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows. Different approaches are described in section \ref{part_of_section_six}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
%
In a nutshell, it must be ensured that the values of an overlap matrix in the first diagonals above and below the main diagonal are at least 0.03 in order to yield reliable results. 
%
There are several other parameters which should be investigated. Basic analysis of the trajectories should be performed to ensure that the ligand does not tumble out of the binding site by measuring its RMSD. Furthermore, distributions of the ligand's dihedral angles can be obtained and compared to experimental values. Thus, poorly sampled conformations can be identified and the issue can be adequately addressed, e.g. by prolonging the simulation time or choosing a different starting conformation. In general, running the same perturbation using different ligand poses may offer insights into whether the calculated free energies are dependent on the initial conformation.
Another easy check can be performed by comparing the results obtained by different free energy estimators (e.g. TI and MBAR). In principal, the outcomes of both estimators should be the same (within a certain range). Hence, if the outcomes deviate by more than a predefined threshold (e.g. 1.0 kcal/mol), the free energies are likely unreliable and further calculations are necessary.

@@

\subsection{Recommendations}
\todo[inline]{ASJSM:Maybe here we have a quick summary on what things should be checked for to make sure simulations are ok?}
\begin{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\item Other considerations for many transformations
\end{itemize}
